---
title: "COVID-19 Team Beta Group Project"
author: "Shree, Andrew, Jack, Oscar, Kelly"
date: 2020-10-18
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

```

## Introduction

The Covid-19 pandemic was undoubtedly one of the most devastating and destructive events any of us have witnessed at a global scale. However, as with any other major crisis, the pandemic didn’t impact every demographic the same. This project attempts to use exploratory data analysis to answer the question, “how did the pandemic and its economic consequences impact men and women differently in the United States?”. 

According to a [policy brief published by the UN](https://www.unwomen.org/-/media/headquarters/attachments/sections/library/publications/2020/policy-brief-the-impact-of-covid-19-on-women-en.pdf?la=en&vs=1406) in April of 2020, women and girls experienced more negative economic consequences, poorer health services (due to the reallocation of medical resources and priorities), and an increase in gender-based violence during the Covid-19 pandemic. An [article published by the McKinsey company](https://www.mckinsey.com/featured-insights/future-of-work/covid-19-and-gender-equality-countering-the-regressive-effects) stated similar statistics, citing that the majority of job losses were experienced by women because of existing gender inequalities. 

Both of these sources looked at how the pandemic impacted people globally. Knowing what happened at a global scale, we were curious about how these trends in gender disparities existed in the United States. Our project uses datasets published by the CDC and Bureau of Labor Statistics to examine how the pandemic impacted men and women differently in the US. 


## Exploratory Data Analysis

First, we load in the data we'll be using. 

```{r}
df_unemploy <- read_csv("data/gender_and_unemployment.csv")
glimpse(df_unemploy)

```

The second figure we wanted to look at was Unemployment rates in the US seperated by gender. 

```{r}
df_unemploy %>%
  ggplot() + 
  geom_line(aes(Date, UnemploymentRate, group = Gender, color = Gender)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  labs(
    y = "Unemployment Rate by Percent", 
    title = "Unemployment rates in the US during the Covid-19 pandemic"
  )
```
**Observations**:
Unemployment rates were roughly the same for men and women before the covid-19 pandemic, around 3%. During the pandemic, both men and women experienced a drastic increase in unemployment rates, however unemployment rates for women were about 3% higher than for men. Around October of 2020, after unemployment rates decreased for both men and women, trends returned to patterns we noticed before the pandemic, where unemployment rates for both men and women are about the same. 


## Final Graphs

## Conclusion

## Sources
