---
title: "COVID-19 Team Beta Group Project"
author: "Shree, Andrew, Jack, Oscar, Kelly"
date: 2020-10-18
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggrepel)
```

## Introduction

The Covid-19 pandemic was undoubtedly one of the most devastating and destructive events any of us have witnessed at a global scale. However, as with any other major crisis, the pandemic didn’t impact every demographic the same. This project attempts to use exploratory data analysis to answer the question, “how did the pandemic and its economic consequences impact men and women differently in the United States?”. 

According to a [policy brief published by the UN](https://www.unwomen.org/-/media/headquarters/attachments/sections/library/publications/2020/policy-brief-the-impact-of-covid-19-on-women-en.pdf?la=en&vs=1406) in April of 2020, women and girls experienced more negative economic consequences, poorer health services (due to the reallocation of medical resources and priorities), and an increase in gender-based violence during the Covid-19 pandemic. An [article published by the McKinsey company](https://www.mckinsey.com/featured-insights/future-of-work/covid-19-and-gender-equality-countering-the-regressive-effects) stated similar statistics, citing that the majority of job losses were experienced by women because of existing gender inequalities. 

Both of these sources looked at how the pandemic impacted people globally. Knowing what happened at a global scale, we were curious about how these trends in gender disparities existed in the United States. Our project uses datasets published by the CDC and Bureau of Labor Statistics to examine how the pandemic impacted men and women differently in the US. 


## Exploratory Data Analysis

First, we load in the data we'll be using. 

```{r}
cases <- read_csv("./data/Provisional_COVID-19_Deaths_by_Sex_and_Age.csv")

df_unemploy <- read_csv("data/gender_and_unemployment.csv")
df_industry <- read_csv("data/DS_Industry_Top_Level_Percent.csv")

#DF_num_employed.csv was a file we generated from the code in BLS_data_manip.Rmd
df_num_employed <- read_csv("data/DF_num_employed.csv")
```

Using a dataset from the CDC, we examine the number of Covid deaths in the US, separated by gender

```{r}
male_data <- cases %>% 
  filter(Sex == "Male", State == "United States", Group == "By Month") %>% 
  mutate(`Start Date` = as.Date(`Start Date`)) %>% 
  mutate(`End Date` = as.Date(`End Date`, format = "%m/%d/%y"))

female_data <- cases %>% 
  filter(Sex == "Female", State == "United States", Group == "By Month") %>% 
  mutate(`Start Date` = as.Date(`Start Date`)) %>% 
  mutate(`End Date` = as.Date(`End Date`, format = "%m/%d/%y"))

```

```{r}
ggplot() +
  geom_smooth(data = male_data, aes(x = `End Date`, y = `COVID-19 Deaths`, color = Sex), se = FALSE) + 
  # geom_point(data = male_data, aes(x = `End Date`, y = `COVID-19 Deaths`), color = "blue") + 
  geom_smooth(data = female_data, aes(x = `End Date`, y = `COVID-19 Deaths`, color = Sex), se = FALSE) + 
  # geom_point(data = female_data, aes(x = `End Date`, y = `COVID-19 Deaths`), color = "red") + 
  ggtitle("Covid Deaths Over Time")

```
*Figure 1*

**Observations**:
From the graph above, we can observe the deaths over time for men and women. Both men and women follow similar trends in number of deaths, with men having one thousand more deaths per day. From April to June, there seems to be a steady decrease in the number of deaths for both men and women. However, there is a drastic increase in the number of deaths for both men and women from the pandemic from October onwards. This phenomenon may be caused by the onset of winter months, where cases frequently increase. 

The second figure we wanted to look at was Unemployment rates in the US separated by gender. 

```{r}
df_unemploy %>%
  ggplot() + 
  geom_line(aes(Date, UnemploymentRate, group = Gender, color = Gender)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  labs(
    y = "Unemployment Rate by Percent", 
    title = "Unemployment rates in the US during the Covid-19 pandemic"
  )
```
*Figure 2*

**Observations**:
Unemployment rates were roughly the same for men and women before the covid-19 pandemic, around 3%. During the pandemic, both men and women experienced a drastic increase in unemployment rates, however unemployment rates for women were about 3% higher than for men. Around October of 2020, after unemployment rates decreased for both men and women, trends returned to patterns we noticed before the pandemic, where unemployment rates for both men and women are about the same. 

In order to better understand why unemployment rates differed so drastically, we determine which industries had predominantly female workers.  


```{r}
OverallPercent <-
  df_industry %>%
  ggplot(aes(Industry, Percent, fill = Percent >= 50)) + 
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.25, hjust=1)) +
  geom_hline(yintercept = 50, linetype = "dashed") 

OverallPercent + ggtitle("Percent of Women Employed in Industry") +
  xlab("Industry") + ylab("Percent")
```
*Figure 3*

**Observations**:
There are only a few industries where women make up 50% or more of the people employed. The four industries that were predominantly female were Education and health services, Leisure and Hospitality, Financial Activities, and Other services. On the other hand, the Construction and Mining industries had the smallest proportion of women. 

Now that we know which industries are predominantly female or male, we look at the percent rate of change of employment for those industries. 

```{r}
df_num_employed %>%
  filter(
    !(Industry %in% c("Nondurable goods", "Total nonfarm", "Service-providing", "Government", "Goods-producing")),
    Group != "All",
    Date > "2018-12-1"
  ) %>%
  ggplot(aes(x = Date, y = Percent_rate_of_change)) + 
  geom_line(aes(color = Group)) +
  facet_wrap(vars(Industry)) +
  scale_color_manual(values = c("blue", "red"))

```
*Figure 4*

**Observations**:
When looking at the individual industry graphs, we can see that very few industries have a negative spike where women do not lose jobs faster than everyone else. In several industries such as Information and Financial Activities the percent rate of change is fairly even. In the Services industries, Retail, and Education, the spike for women is much worse than that of their other coworkers. This visualization also gives us a sense of which industries were most impacted by the Covid-19 pandemic as far as employment goes. 



## Final Graphs
```{r}
theme_common <- function() {
  theme_minimal() %+replace%
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(margin = margin(4, 4, 4, 4), size = 16),
    axis.title.y = element_text(margin = margin(4, 4, 4, 4), size = 16, angle = 90),

    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),

    strip.text.x = element_text(size = 12),
    strip.text.y = element_text(size = 12),

    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_line(color = "grey90"),

    aspect.ratio = 4 / 4,

    plot.margin = unit(c(t = +0, b = +0, r = +0, l = +0), "cm"),
    plot.title = element_text(size = 18),
    plot.title.position = "plot",
    plot.subtitle = element_text(size = 16),
    plot.caption = element_text(size = 12)
  )
}
```


```{r}
df_num_employed %>%
  filter(
    Industry == "Other services",
    Group != "All",
    Date > "2018-12-1"
  ) %>%
  ggplot(aes(x = Date, y = Percent_rate_of_change)) + 
  geom_line(aes(color = Group)) + 
  scale_color_manual(labels = c("Other", "Women"), values = c("blue", "red")) + 
  theme_common() + 
  labs(
    y = "Percent rate of change of employed",
    x = "Date",
    title = "United States Services Industry",
    color = "Gender"

  )

df_num_employed%>%
  filter(
    Industry == "Education and health services",
    Group != "All",
    Date > "2018-12-1"
  ) %>%
  ggplot(aes(x = Date, y = Percent_rate_of_change)) + 
  geom_line(aes(color = Group)) + 
  scale_color_manual(labels = c("Other", "Women"), values = c("blue", "red")) + 
  theme_common() + 
  labs(
    y = "Percent rate of change of employed",
    x = "Date",
    title = "Education and Health Services Industry in the United States",
    color = "Gender"

  )

df_num_employed %>%
  filter(
    Industry == "Total nonfarm",
    Group != "All",
    Date > "2018-12-1"
  ) %>%
  ggplot(aes(x = Date, y = Percent_rate_of_change)) + 
  geom_line(aes(color = Group)) + 
  scale_color_manual(labels = c("Other", "Women"), values = c("blue", "red")) +
  theme_common() +
  labs(
    y = "Percent rate of change of employed",
    x = "Date",
    title = "Overall Employment in the United States",
    color = "Gender"
  )
```

**Observations**
In the two industries that were most dominated by women, "Other services" and "Education and health services", women lost their jobs at a higher  rate than men even when accounting for gender imbalance in the industry. Looking at the min and max peaks of these graphs, we can deduce that after the peak of the pandemic when job losses were greatest, increase in employment did not completely offset the number of people who lost their jobs. We know this is true because looking at figure 2, unemployment rates are still higher than pre-pandemic levels.  

## Conclusion

The figure 1 we examined answers how the pandemic impacted men and women differently in terms of Covid deaths. When looking at the number of Covid death in the US, we see that more men died of Covid than women. After seeing this data we pivoted our question to focus more on the economic impacts of Covid on different genders rather than just the direct physical impacts of the pandemic. Our second EDA graph revealed that while unemployment rates spiked for both  men and women in the US, women experienced a noticeably more drastic increase in unemployment rates than men (figure 2). 

To begin to understand why that might be, we looked at each industry that was categorized by the Bureau of Labor Statistics, and determined which ones were predominantly women (figure 3). Next, we looked at the percent change in employment rates for women and men in each of those industries (figure 4). 

For our final graphs, we examined the "Education and health services" industry and "Other services" industry because they were the two industries with the highest proportion of women. In both of these industries, women were let go at a higher percent rate of change than men. These two industries also experienced more job losses than most other industries. In other words, not only did predominantly female industries experience more job losses than other predominantly male industries, but within those female dominated industries, women were let go at a higher rate than men. 

Unsurprisingly, our findings corroborate those that were stated in the UN report and McKinsey article. It's a commonly recognized reality that events like a pandemic serve to exacerbate existing inequalities, whether that's in regards to race, gender, or wealth. In going through these processes of retrieving raw data, shaping it into usable information, and visualizing it to examine patterns and outliers, we hope to gain a better understanding of how data can be used to educate us about ways in which our systems can be improved. 

## Sources

“CDC Covid Data Tracker.” Centers for Disease Control and Prevention, Centers for Disease Control and         
\ \ \ \ \ \ Prevention, https://covid.cdc.gov/covid-data-tracker/#demographics. 

“Databases, Tables &amp; Calculators by Subject.” U.S. Bureau of Labor Statistics, U.S. Bureau of Labor Statistics, 
\ \ \ \ \ \ https://www.bls.gov/data/#unemployment. 

Madgavkar, Anu, et al. “Covid-19 and Gender Equality: Countering the Regressive Effects.” McKinsey &amp; Company, 
\ \ \ \ \ \ McKinsey &amp; Company, 10 Apr. 2021, https://www.mckinsey.com/featured-insights/future-of-work/covid-19-and-
\ \ \ \ \ \ gender-equality-countering-the-regressive-effects. 

“Www.unwomen.org.” UNWomen.org, United Nations, 9 Apr. 2020, 
\ \ \ \ \ \ https://www.unwomen.org/-/media/headquarters/attachments/sections/library/publications/2021/egm-background
\ \ \ \ \ \ -papers-financing-for-gender-equality-in-the-hiv-response-en.pdf?la=en&amp;vs=5214. 