---
title: "COVID-19 Team Beta Group Project"
author: "Shree, Andrew, Jack, Oscar, Kelly"
date: 2020-10-18
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggrepel)
```

## Introduction

The Covid-19 pandemic was undoubtedly one of the most devastating and destructive events any of us have witnessed at a global scale. However, as with any other major crisis, the pandemic didn’t impact every demographic the same. This project attempts to use exploratory data analysis to answer the question, “how did the pandemic and its economic consequences impact men and women differently in the United States?”. 

According to a [policy brief published by the UN](https://www.unwomen.org/-/media/headquarters/attachments/sections/library/publications/2020/policy-brief-the-impact-of-covid-19-on-women-en.pdf?la=en&vs=1406) in April of 2020, women and girls experienced more negative economic consequences, poorer health services (due to the reallocation of medical resources and priorities), and an increase in gender-based violence during the Covid-19 pandemic. An [article published by the McKinsey company](https://www.mckinsey.com/featured-insights/future-of-work/covid-19-and-gender-equality-countering-the-regressive-effects) stated similar statistics, citing that the majority of job losses were experienced by women because of existing gender inequalities. 

Both of these sources looked at how the pandemic impacted people globally. Knowing what happened at a global scale, we were curious about how these trends in gender disparities existed in the United States. Our project uses datasets published by the CDC and Bureau of Labor Statistics to examine how the pandemic impacted men and women differently in the US. 


## Exploratory Data Analysis

First, we load in the data we'll be using. 

```{r}
cases <- read_csv("./data/Provisional_COVID-19_Deaths_by_Sex_and_Age.csv")

df_unemploy <- read_csv("data/gender_and_unemployment.csv")
df_num_employed <- read_csv("data/DF_num_employed.csv")
df_industry <- read_csv("data/DS_Industry_Top_Level_Percent.csv")

```

```{r}
male_data <- cases %>% 
  filter(Sex == "Male", State == "United States", Group == "By Month") %>% 
  mutate(`Start Date` = as.Date(`Start Date`)) %>% 
  mutate(`End Date` = as.Date(`End Date`, format = "%m/%d/%y"))

female_data <- cases %>% 
  filter(Sex == "Female", State == "United States", Group == "By Month") %>% 
  mutate(`Start Date` = as.Date(`Start Date`)) %>% 
  mutate(`End Date` = as.Date(`End Date`, format = "%m/%d/%y"))


```

```{r}
ggplot() +
  geom_smooth(data = male_data, aes(x = `End Date`, y = `COVID-19 Deaths`, color = Sex), se = FALSE) + 
  # geom_point(data = male_data, aes(x = `End Date`, y = `COVID-19 Deaths`), color = "blue") + 
  geom_smooth(data = female_data, aes(x = `End Date`, y = `COVID-19 Deaths`, color = Sex), se = FALSE) + 
  # geom_point(data = female_data, aes(x = `End Date`, y = `COVID-19 Deaths`), color = "red") + 
  ggtitle("Covid Deaths Over Time")

```

From the graph above, we can observe the deaths over time for men and women. Both men and women follow similar trends in number of deaths, with men having one thousand more deaths per day. From April to June, there seems to be a steady decrease in the number of deaths for both men and women. However, there is a drastic increase in the number of deaths for both men and women from the pandemic from October onwards. This phenomenon may be caused by the onset of winter months, where cases frequently increase. 

The second figure we wanted to look at was Unemployment rates in the US separated by gender. 

```{r}
df_unemploy %>%
  ggplot() + 
  geom_line(aes(Date, UnemploymentRate, group = Gender, color = Gender)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  labs(
    y = "Unemployment Rate by Percent", 
    title = "Unemployment rates in the US during the Covid-19 pandemic"
  )
```
**Observations**:
Unemployment rates were roughly the same for men and women before the covid-19 pandemic, around 3%. During the pandemic, both men and women experienced a drastic increase in unemployment rates, however unemployment rates for women were about 3% higher than for men. Around October of 2020, after unemployment rates decreased for both men and women, trends returned to patterns we noticed before the pandemic, where unemployment rates for both men and women are about the same. 

```{r}
OverallPercent <-
  df_industry %>%
  ggplot(aes(Industry, Percent, fill = Percent >= 50)) + 
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.25, hjust=1)) +
  geom_hline(yintercept = 50, linetype = "dashed") 

OverallPercent + ggtitle("Percent of Women Employed in Industry") +
  xlab("Industry") + ylab("Percent")
```
**Observations**:
There are only a few industries where women make up 50% of the people employed. Additionally there is only one industry that could be considered as female dominted which is Education and Health Services. Education and Health Services is also an industry which is essential and ideally should have experienced the smallest impact on employment. Given how many women work in this indutry, it's interesting that the unemployment rate for women was noticeably higher than men. 

## Final Graphs
```{r}
theme_common <- function() {
  theme_minimal() %+replace%
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(margin = margin(4, 4, 4, 4), size = 16),
    axis.title.y = element_text(margin = margin(4, 4, 4, 4), size = 16, angle = 90),

    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),

    strip.text.x = element_text(size = 12),
    strip.text.y = element_text(size = 12),

    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_line(color = "grey90"),

    aspect.ratio = 4 / 4,

    plot.margin = unit(c(t = +0, b = +0, r = +0, l = +0), "cm"),
    plot.title = element_text(size = 18),
    plot.title.position = "plot",
    plot.subtitle = element_text(size = 16),
    plot.caption = element_text(size = 12)
  )
}
```

```{r}
filename <- "./data/DS_monthly_breakdowns_all.csv"
df_all <- read_csv(filename)

filename <- "./data/DS_monthly_breakdowns_women.csv"
df_women <- read_csv(filename)
```

```{r}
colnames(df_women)[4] <- "Count_Women"

colnames(df_all)[4] <- "Count_All"

```


```{r}
df_women <-
  df_women %>%
  mutate(
    Date = as.Date(paste(Year, Month, 1, sep = " "), format = '%Y %b %d')
  )

df_all <-
  df_all %>%
  mutate(
    Date = as.Date(paste(Year, Month, 1, sep = " "), format = '%Y %b %d')
  )
```


```{r}
df_all_vs_women <-
  df_all %>%
  left_join(df_women) %>%
  subset(select = c(3,4,5,6))
```

```{r}
df_all_vs_women <-
  df_all_vs_women %>%
  mutate(
    Count_notWomen = Count_All - Count_Women
  ) %>%
  pivot_longer(
    names_to = c(".value", "Group"),
    names_sep = "_",
    cols = c(Count_All, Count_Women, Count_notWomen)
  )
```


```{r}
df_all_vs_women <-
  df_all_vs_women %>%
  group_by(Industry, Group) %>%
  arrange(Industry, Group, Date) %>%
  mutate(
    Percent_rate_of_change =  100 * (Count - lag(Count))/lag(Count)
  ) %>%
  ungroup()

write.csv(df_all_vs_women, ".\\data\\DF_num_employed.csv")
```

```{r}
df_all_vs_women %>%
  filter(
    Industry == "Service-providing",
    Group != "All",
    Date > "2018-12-1"
  ) %>%
  ggplot(aes(x = Date, y = Percent_rate_of_change)) + 
  geom_line(aes(color = Group)) + 
  facet_wrap(vars(Industry)) + 
  scale_color_manual(values = c("blue", "red")) + 
  theme_common()

df_all_vs_women %>%
  filter(
    Industry == "Education and health services",
    Group != "All",
    Date > "2018-12-1"
  ) %>%
  ggplot(aes(x = Date, y = Percent_rate_of_change)) + 
  geom_line(aes(color = Group)) + 
  facet_wrap(vars(Industry)) + 
  scale_color_manual(values = c("blue", "red")) + 
  theme_common()

df_all_vs_women %>%
  filter(
    Industry == "Goods-producing",
    Group != "All",
    Date > "2018-12-1"
  ) %>%
  ggplot(aes(x = Date, y = Percent_rate_of_change)) + 
  geom_line(aes(color = Group)) + 
  facet_wrap(vars(Industry)) + 
  scale_color_manual(values = c("blue", "red")) + 
  theme_common()

df_all_vs_women %>%
  filter(
    Industry == "Total nonfarm",
    Group != "All",
    Date > "2018-12-1"
  ) %>%
  ggplot(aes(x = Date, y = Percent_rate_of_change)) + 
  geom_line(aes(color = Group)) + 
  facet_wrap(vars(Industry)) + 
  scale_color_manual(values = c("blue", "red")) +
  theme_common()
```
**Observations**
The first set of data we looked at was the rate at which jobs were lost for women and others in sectors of industry. We chose to look as the rate of job loss by gender to account for the uneven distribution of employment in many of these industries. In most industries, women job loss was notably higher than non-women job loss but there were a few industries such as goods producing where that trend is reversed. Despite this, we can see from the total nonfarm graph that women lost their jobs proportionately more than others. 


## Conclusion

When looking at the number of Covid death in the US, we see that more men died of Covid than women. After seeing this data we pivoted our question to focus more on the economic impacts of Covid on different genders rather than just the direct physical impacts of the pandemic. Our second EDA graph revealed that while unemployment rates spiked for both  men and women in the US, women experienced a noticeably more drastic increase in unemployment rates than men. To begin to understand why that might be, we looked at each industry that was categorized by the Bureau of Labor Statistics, and determined which ones were predominantly women. Next, we looked at the percent change in employment rates for women and men in each of those industries. For our final graphs, we examined the 


Looking at our final graphs, it's clear that unemployment rates increased for both men and women. However, we were able to see the trends described in the UN report and McKinsey paper-- women experienced a more drastic increase in unemployment rates compared to men during the first few months of the pandemic. Throughout the pandemic, industries that were predominantly women, such as the Education and 
The unemployment rates increased for both men and women and there are a few industries which are either equal employers or female dominated but we can see that regardless of initial gender ratio in industries, the percent rate of change for women's unemployment was significantly more therefore more women were impacted by the pandemic where it came to employment and job losses.

## Sources
